04-analisis

### 2024-05-27 03-methods.qmd
Arreglos presentación modelos

## Data, variables and method

### Data

The First Study of Civic Education in Chile, conducted by the Agency for Quality Education of the Ministry of Education in 2017, serves as the main data source. This database is composed by a civic knowledge test score and a series of items batteries measuring different aspects of citizenship. The target population includes 8th-grade students in 242 schools nationwide, as well as their parents and teachers. Initially, the database contains 8,589 student observations and 6,770 parent observations. To ensure higher data quality and considering the survey's unit of analysis, we removed 171 student cases and 79 parent cases that exhibited repetitive and careless response patterns (Gottfried et al., 2022). Additionally, we utilized data from the System for Measuring the Quality of Education (SIMCE) conducted by the Ministry of Education in 2017, which provides school-level information such as administrative dependence, socioeconomic classification, and results obtained in standardized mathematics and language tests. After processing the variables and removing missing cases, the final database for this study includes a two-level stratified sample composed of 5,047 students and parents (level 1) nested within 231 schools (level 2).

### Variables

@@

**Justification of Inequality**: The dependent variables in this study are market justice preferences. This construct is measured through three variables that address the degree of justification regarding whether access to social services in pensions, education, and health should be income conditional. Justification of inequality in health is measured by the item: "Is it fair in Chile that people with higher incomes can access better healthcare than those with lower incomes?" The same question is asked for education and pensions. In all cases, respondents indicate their preferences on a Likert scale ranging from "completely disagree" (1) to "completely agree" (4). Additionally, we include a summarized indicator of "market justice preferences", measured by an average index across these items (α = 0.86), with values ranging from 1 to 4, where higher values represent greater preferences for market justice (see @tbl-desc-dependientes). We analyzed these items independently as well as by the average index.

@@

@tbl-desc-independent shows the independent variables used, their response categories and their frequencies.

@@

```{r, results='asis', echo=FALSE}
#| tbl-cap: "Independent variables"
#| label: tbl-desc-independent

@@

This study focuses on two school-level characteristics: socioeconomic status and academic performance. Socioeconomic status is assessed using the Ministry of Education's classification, measured as an ordinal item with five categories ranging from "low" (1) to "high" (5). Academic performance is measured using the school's results in the SIMCE (System of Measurement of Educational Quality) standardized tests, which are administered yearly at different educational levels in areas such as language and mathematics. These results are categorized as "low," "medium," and "high." The items used, response categories, and their frequencies are detailed in @tbl-desc-school.

@@

df<-dfSummary(school[,c(2,4)],
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

>>>>

# Data, variables and method

## Data

The First Study of Civic Education in Chile, conducted by the Agency for Quality Education of the Ministry of Education in 2017, serves as the main data source. This database is composed by a civic knowledge test score and a series of items batteries measuring different aspects of citizenship. The target population includes 8th-grade students in 242 schools nationwide, as well as their parents and teachers. Initially, the database contains 8,589 student observations and 6,770 parent observations. To ensure higher data quality and considering the survey's unit of analysis, we removed 171 student cases and 79 parent cases that exhibited repetitive and careless response patterns [@gottfried_autocorrelation_2022]. Additionally, we utilized data from the System for Measuring the Quality of Education (SIMCE) conducted by the Ministry of Education in 2017, which provides school-level information such as administrative dependence, socioeconomic classification, and results obtained in standardized mathematics and language tests. After processing the variables and removing missing cases, the final database for this study includes a two-level stratified sample composed of 5,047 students and parents (level 1) nested within 231 schools (level 2).

## Variables

@@

**Market Justice Preferences**: The dependent variables in this study are market justice preferences. This construct is measured through three variables that address the degree of justification regarding whether access to social services in pensions, education, and health should be income conditional. Justification of inequality in health is measured by the item: "Is it fair in Chile that people with higher incomes can access better healthcare than those with lower incomes?" The same question is asked for education and pensions. In all cases, respondents indicate their preferences on a Likert scale ranging from "completely disagree" (1) to "completely agree" (4). Additionally, we include a summarized indicator of "market justice preferences", measured by an average index across these items (α = 0.86), with values ranging from 1 to 4, where higher values represent greater preferences for market justice (see @tbl-desc-dependientes). We analyzed these items independently as well as by the average index.

@@

@tbl-desc-independent shows the individual level variables used, their response categories and their frequencies.

@@

```{r, results='asis', echo=FALSE}
#| tbl-cap: "Individual level variables"
#| label: tbl-desc-independent

@@

This study focuses on two school-level characteristics: socioeconomic status and academic performance. Socioeconomic status is assessed using the Ministry of Education's classification, measured as an ordinal item with five categories ranging from "low" (1) to "high" (5). Academic performance is measured using the school's results in the SIMCE (System of Measurement of Educational Quality) standardized tests, which are administered yearly at different educational levels in areas such as language and mathematics. These results are categorized as "low," "medium," and "high." The contextual level items used, response categories, and their frequencies are detailed in @tbl-desc-school.

@@

df<-dfSummary(school,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

### 2024-05-27 04-analisis.qmd
Arreglos presentación modelos

## Analysis

### Descriptive analysis

@@

Likert plots

@@

### Multivariate results

The results of the cumulative link mixed models for the ordinal dependent variables of justice in differential access to education, health, and pensions by income are presented in @tbl-ordinal-reg. This table shows the effects of individual and contextual-level independent variables related to our hypothesis; however, the models also incorporate the control variables utilized in the study for each of the three dependent variables separately (complete models are available in the [Appendix](#appendix)).

@@

ccoef <- list(
  inteligencia_esc = "School talent",
  esfuerzo_esc = "School effort",
  inteligencia_soc = "Social talent",
  esfuerzo_soc = "Social effort",
  merito_soc = "Deservingness",
  "educacion_recEd Media" = "Secondary",
  "educacion_recEd Tecnica" = "Higher tec.",
  "educacion_recUniversidad o posgrado" = "University or posgraduate",
  "educacion_recNs/Nr" = "Missing",
  mean_educ = "Prop. university level at school",
  "cod_grupoMedio bajo" = "SES Medium low",
  cod_grupoMedio = "SES Medium",
  "cod_grupoMedio alto" = "SES Medium high",
  cod_grupoAlto = "SES High",
  simceMedio = "Simce Medium",
  simceAlto = "Simce High"
)

texreg::htmlreg(l = list(modelos_odd_educ[[2]],
                         modelos_odd_educ[[3]],
                         modelos_odd_health[[2]],
                         modelos_odd_health[[3]],
                         modelos_odd_pension[[2]],
                         modelos_odd_pension[[3]]),   
                  stars = c(0.05, 0.01, 0.001),
                  custom.note = "Note: Coefficients models in odds-ratio. Control variables are included. Standard errors in parentheses. %stars",
                  custom.header = list("Education" = 1:2, "Health" = 3:4, "Pensions" = 5:6),
                  custom.model.names = rep(c("Model 1", "Model 2"), 3),
                  custom.coef.map = ccoef,
                  leading.zero = T,
                  scalebox = 0.60,
                  groups = list("Parental education (Ref.= 8th grade or less)" = 6:9,
                                "Socioeconomic level (Ref.= Low)" = 11:14,
                                "Achievement score (Ref.= Low)" = 15:16),
                 custom.gof.rows = list("Controls"=rep("Yes",6)))

@@

reg0_bienestar <- lmer(bienestar ~ 1 + (1 | mrbd), data=data_bienestar)
reg1_bienestar <- lmer(bienestar ~ 1 + inteligencia_esc + esfuerzo_esc + inteligencia_soc + esfuerzo_soc + merito_soc + (1 | mrbd), data=data_bienestar)
reg2_bienestar <- lmer(bienestar ~ 1 + inteligencia_esc + esfuerzo_esc + inteligencia_soc + esfuerzo_soc + merito_soc + educacion_rec + libros_rec + acc_tec + internet_rec + (1 | mrbd), data=data_bienestar)
reg3_bienestar <- lmer(bienestar ~ 1 + mean_educ + cod_depe2 + cod_grupo + simce + (1 | mrbd), data=data_bienestar)
reg4_bienestar <- lmer(bienestar ~ 1 + esfuerzo_soc + merito_soc + inteligencia_soc + esfuerzo_esc + inteligencia_esc + educacion_rec + libros_rec + acc_tec + internet_rec + mean_educ + cod_depe2 + cod_grupo + simce + (1 | mrbd), data=data_bienestar)

test01<- anova(reg0_bienestar,reg1_bienestar,test = "Chisq")
test02<- anova(reg0_bienestar,reg2_bienestar,test = "Chisq")
test03<- anova(reg0_bienestar,reg3_bienestar,test = "Chisq")
test04<- anova(reg0_bienestar,reg4_bienestar,test = "Chisq")

test.pvalues1<- test01$`Pr(>Chisq)`[2]
test.pvalues2<- test02$`Pr(>Chisq)`[2]
test.pvalues3<- test03$`Pr(>Chisq)`[2]
test.pvalues4<- test04$`Pr(>Chisq)`[2]

deviance1 <- test01$deviance[2]
deviance2 <- test02$deviance[2]
deviance3 <- test03$deviance[2]
deviance4 <- test04$deviance[2]
```

@@

texreg::htmlreg(list(reg1_bienestar, reg2_bienestar),
        custom.model.names = c("Model 1",
                               "Model 2"),
        custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
        omit.coef = "(internet_rec)",
        caption = "",
        custom.coef.names = c("Intercept",  
                              "School talent",
                              "School effort", 
                              "Social talent",  
                              "Social effort",
                              "Deservingness",
                              "Secondary", 
                              "Higher tec.", 
                              "University or posgraduate", 
                              "Missing", 
                              "More than 25 books (Ref. Less than 25)", 
                              "Technology access"),
        groups = list("Parental education (Ref.= 8th grade or less)"=7:10),
        custom.gof.rows=list("Deviance" = c(deviance1,
                                            deviance2),
                             "Deviance Test (p)" = c(test.pvalues1,
                                                       test.pvalues2))
        )

@@

```{r echo=FALSE, results='asis'}
#| tbl-cap: "Contextual effects of multilevel regression models"
#| label: tbl-contextual-reg

texreg::htmlreg(list(reg3_bienestar, reg4_bienestar),
        custom.model.names = c("Model 1",
                               "Model 2"),
        custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
        omit.coef = c("inteligencia_esc|esfuerzo_esc|inteligencia_soc|esfuerzo_soc|merito_soc|educacion_rec|libros_rec|acc_tec|internet_rec"),
        caption = "",
        custom.coef.names = c("Intercept", 
                              "Prop. university level at school",
                              "Subsidized private", 
                              "Private", 
                              "SES Medium low", 
                              "SES Medium", 
                              "SES Medium high", 
                              "SES High", 
                              "Simce Medium", 
                              "Simce High"),
        groups = list("Administrative dependence (Ref.= Public)"=3:4,
                      "Socioeconomic level (Ref.= Low)"=5:8,
                      "Achievement score (Ref.= Low)"=9:10),
        custom.gof.rows=list("Deviance" = c(deviance3,
                                            deviance4),
                             "Deviance Test (p)" = c(test.pvalues3,
                                                     test.pvalues4))
        )

>>>>

# Analysis

## Descriptive analysis

@@

@fig-marketjustice shows the frequency distribution for the three market justice preference items (pensions, education, and health). The majority of respondents disagree (63.0% in total) with the idea that it is just for higher-income individuals to access better pensions. However, a notable portion (27.0%) agrees, and a smaller minority (10.1%) strongly agrees. Similar to pensions, a majority of respondents (58.8%) disagree with the idea that higher-income individuals should have access to better education. However, a considerable portion (28.6%) agrees, and a slightly larger minority (12.6%) strongly agrees compared to pensions. In the context of health, an even larger majority of respondents (66.4%) disagree with the notion that higher-income individuals should access better healthcare services. A smaller percentage (23.6%) agrees, and a minority (10.0%) strongly agrees. Overall, @fig-marketjustice shows a predominant sentiment against the justice of higher-income individuals accessing better services across all three areas. The strongest opposition is observed in the health sector. Despite the overall disagreement, a significant portion of respondents agree with the statement, particularly in education, where agreement and strong agreement are slightly higher compared to pensions and health.

```{r echo=FALSE}
#| fig-cap: "Market justice preferences in education, health and pensions"
#| fig-asp: 1.22
#| label: fig-marketjustice

data_rec %>% 
  select(just_salud, just_pension, just_educ) %>% 
  sjPlot::plot_stackfrq(expand.grid =TRUE,
                      geom.colors = c("#7d7d7d","#ababab","#cecece","#ebebeb"),
                      show.total = FALSE, 
                      legend.labels = c("Strongly disagree", "Disagree", "Agree", "Strongly agree"),
                      axis.labels = c("Health", "Pensions", "Education")) +
 # theme_light(base_size = 16) +
  theme(legend.position="bottom", text = element_text(size = 14),
        plot.caption = element_text(hjust = 1))
```

@fig-meritocracy shows the frequency distribution for the five meritocracy items for the dimensions of school and social meritocracy. While a significant portion of respondents (29.2%) strongly agrees that talent is important to get good grades, a larger majority (78.6% combining agree and strongly agree) supports this notion. However, the strong agreement is not as pronounced as for school effort. An overwhelming majority (73.1%) agree that students effort is important to get good grades, which is a much higher percentage compared to the strong agreement for talent (29.2%). This indicates that people believe effort is more important than talent at school. There is a strong belief that both effort and talent are important, with a particularly high agreement for effort (73.1% strongly agree).

```{r echo=FALSE}
#| fig-cap: "Social and school meritocracy"
#| fig-asp: 1.22
#| label: fig-meritocracy

plot<-data_rec %>% 
  select(merito_soc, esfuerzo_soc, inteligencia_soc, esfuerzo_esc, inteligencia_esc) %>% 
  sjPlot::plot_stackfrq(expand.grid =TRUE,
                      geom.colors = c("#7d7d7d","#ababab","#cecece","#ebebeb"),
                      show.total = FALSE, 
                      vjust=c(rep(c("center"),12), "bottom", "top", rep(c("center"), 6)),
                      legend.labels = c("Strongly disagree", "Disagree", "Agree", "Strongly agree"),
                      axis.labels = c("Deservingness", "Social effort", "Social talent", "School effort", "School talent")) +
  #theme_light(base_size = 16) +
  theme(legend.position="bottom", text = element_text(size = 14),
        plot.caption = element_text(hjust = 1))

plot + 
  annotate("text", x = 5.5, y = 0, label = "School meritocracy", size = 5, hjust = 0, vjust = 1) +
  annotate("text", x = 3.5, y = 0, label = "Social meritocracy", size = 5, hjust = 0, vjust = 1)
```

In relation with social meritocracy, @fig-meritocracy shows that a majority of respondents (63.7% combining agree and strongly agree) agree that people are rewarded for their talent in society, with a notable portion (17.1%) strongly agreeing. A majority (62.0% combining agree and strongly agree) agree that people are rewarded for their effort in society. Similar to social talent, the perception is positive but less overwhelming than school meritocracy. The perception of whether people get what they deserve in society is more divided. While a slim majority (56.5% combining agree and strongly agree) agree, a significant minority (43.6%) disagrees, showing that this is the most contentious item among respondents. There is general agreement that people are rewarded for talent and effort in society, but the consensus is weaker compared to school meritocracy. Deservingness in society is the most contentious, with significant disagreement among respondents.

@@

## Multivariate results

The results of the cumulative link mixed models for the ordinal dependent variables of justice in differential access to health, pensions and education by income are presented in @tbl-ordinal-reg. This table shows the effects of individual and contextual-level independent variables related to our hypothesis; however, the models also incorporate the control variables utilized in the study for each of the three dependent variables separately (complete models are available in the [Appendix](#appendix)).

@@

ccoef <- list(
  inteligencia_esc = "School talent",
  esfuerzo_esc = "School effort",
  inteligencia_soc = "Social talent",
  esfuerzo_soc = "Social effort",
  merito_soc = "Deservingness",
  "educacion_recEd Media" = "Secondary",
  "educacion_recEd Tecnica" = "Higher tec.",
  "educacion_recUniversidad o posgrado" = "University or posgraduate",
  "educacion_recNs/Nr" = "Missing",
  "cod_grupoMedio bajo" = "SES Medium low",
  cod_grupoMedio = "SES Medium",
  "cod_grupoMedio alto" = "SES Medium high",
  cod_grupoAlto = "SES High",
  simceMedio = "Simce Medium",
  simceAlto = "Simce High"
)

texreg::htmlreg(l = list(modelos_odd_health[[2]],
                         modelos_odd_health[[3]],
                         modelos_odd_pension[[2]],
                         modelos_odd_pension[[3]],
                         modelos_odd_educ[[2]],
                         modelos_odd_educ[[3]]),   
                caption = NULL,
                stars = c(0.05, 0.01, 0.001),
                custom.note = "Note: Coefficients models in odds-ratio. Control variables are included. Standard errors in parentheses. %stars",
                custom.header = list("Health" = 1:2, "Pensions" = 3:4, "Education" = 5:6),
                custom.model.names = rep(c("Model 1", "Model 2"), 3),
                custom.coef.map = ccoef,
                leading.zero = T,
                scalebox = 0.60,
                groups = list("Parental education (Ref.= 8th grade or less)" = 6:9,
                                "Socioeconomic level (Ref.= Low)" = 10:13,
                                "Achievement score (Ref.= Low)" = 14:15),
                custom.gof.rows = list("Controls"=rep("Yes",6)),
                include.loglik = FALSE, include.aic = FALSE)

@@

reg0_bienestar <- lmer(bienestar ~ 1 + (1 | mrbd), data = data_bienestar)

performance::icc(reg0_bienestar, by_group = T)

reg1_bienestar <- lmer(bienestar ~ 1 + inteligencia_esc + esfuerzo_esc + 
                         inteligencia_soc + esfuerzo_soc + merito_soc + 
                         acc_tec + internet_rec +
                         (1 | mrbd), data=data_bienestar)

reg2_bienestar <- lmer(bienestar ~ 1 + inteligencia_esc + esfuerzo_esc + 
                         inteligencia_soc + esfuerzo_soc + merito_soc + 
                         educacion_rec + libros_rec + acc_tec + internet_rec +
                         (1 | mrbd), data=data_bienestar)

reg3_bienestar <- lmer(bienestar ~ 1 + mean_educ + cod_depe2 + cod_grupo +
                         simce + (1 | mrbd), data=data_bienestar)

reg4_bienestar <- lmer(bienestar ~ 1 + esfuerzo_soc + merito_soc + 
                         inteligencia_soc + esfuerzo_esc + inteligencia_esc + 
                         educacion_rec + libros_rec + acc_tec + internet_rec + 
                         mean_educ + cod_depe2 + cod_grupo + 
                         (1 | mrbd), data=data_bienestar)

reg5_bienestar <- lmer(bienestar ~ 1 + esfuerzo_soc + merito_soc + 
                         inteligencia_soc + esfuerzo_esc + inteligencia_esc + 
                         educacion_rec + libros_rec + acc_tec + internet_rec + 
                         mean_educ + cod_depe2 + cod_grupo + simce + 
                         (1 | mrbd), data=data_bienestar)

test01 <- anova(reg0_bienestar, reg1_bienestar, test = "Chisq")
test02 <- anova(reg0_bienestar, reg2_bienestar, test = "Chisq")
test03 <- anova(reg0_bienestar, reg3_bienestar, test = "Chisq")
test04 <- anova(reg0_bienestar, reg4_bienestar, test = "Chisq")
test05 <- anova(reg0_bienestar, reg5_bienestar, test = "Chisq")

test.pvalues1 <- test01$`Pr(>Chisq)`[2]
test.pvalues2 <- test02$`Pr(>Chisq)`[2]
test.pvalues3 <- test03$`Pr(>Chisq)`[2]
test.pvalues4 <- test04$`Pr(>Chisq)`[2]
test.pvalues5 <- test05$`Pr(>Chisq)`[2]

deviance1 <- test01$deviance[2]
deviance2 <- test02$deviance[2]
deviance3 <- test03$deviance[2]
deviance4 <- test04$deviance[2]
deviance5 <- test05$deviance[2]
```

@@

ccoef <- list(
  "(Intercept)" = "Intercept",
  inteligencia_esc = "School talent",
  esfuerzo_esc = "School effort",
  inteligencia_soc = "Social talent",
  esfuerzo_soc = "Social effort",
  merito_soc = "Deservingness",
  "educacion_recEd Media" = "Secondary",
  "educacion_recEd Tecnica" = "Higher tec.",
  "educacion_recUniversidad o posgrado" = "University or posgraduate",
  "educacion_recNs/Nr" = "Missing",
  "cod_grupoMedio bajo" = "SES Medium low",
  cod_grupoMedio = "SES Medium",
  "cod_grupoMedio alto" = "SES Medium high",
  cod_grupoAlto = "SES High",
  simceMedio = "Simce Medium",
  simceAlto = "Simce High"
)

texreg::htmlreg(list(reg1_bienestar, reg2_bienestar, reg4_bienestar, reg5_bienestar),
                custom.model.names = c("Model 1",
                                       "Model 2",
                                       "Model 3",
                                       "Model 4"),
                caption = NULL,
                stars = c(0.05, 0.01, 0.001),
                custom.note = "%stars",
                custom.coef.map = ccoef,
                leading.zero = T,
                scalebox = 0.60,
                groups = list("Parental education (Ref.= 8th grade or less)" = 7:10,
                              "Socioeconomic level (Ref.= Low)" = 11:14,
                              "Achievement score (Ref.= Low)" = 15:16),
                custom.gof.rows = list("Controls"=rep("Yes",4),
                                       "Deviance" = c(deviance1,
                                                    deviance2,
                                                    deviance4,
                                                    deviance5),
                                       "Deviance Test (p)" = c(test.pvalues1,
                                                             test.pvalues2,
                                                             test.pvalues4,
                                                             test.pvalues5)),
                include.loglik = FALSE,include.aic = FALSE)

@@
